---
title: "Lab 07 - PCA"
author: "[Your Name Here]"
date: "Due 07 November at 11pm"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
---

This lab is based on material from Chapter 15 of Prof. Cosma Shalizi's
book [_Advanced Data Analysis from an Elementary Point of
View_](https://www.stat.cmu.edu/~cshalizi/ADAfaEPoV/).

```{r setup, include=FALSE}
library(tidyverse)
library(Stat406)
theme_set(theme_bw())
```

In the era before large language models,
information retrieval systems (like search engines) and people doing
computational text analysis often represented documents as what are called **bags
of words**: documents are represented as vectors, where each component counts
how many times each word in the dictionary appears in the text. This throws away
information about word order, but gives us something we can work with
mathematically. Part of the representation of one document might look like:

```{r}
data.frame(nyt_raw[1:6, 2:11])
```

and so on through to "zebra", "zoology", "zygote", etc. to the end of the
dictionary. These vectors are very, very large! At least in English and similar
languages, these bag-of-word vectors have three outstanding properties:

1. most words don't appear in most documents, so these representations are
**sparse** containing many 0's.
1. a few words appear often in most documents (like "a", "an", "the", "of",
etc.). These don't provide much information about the content of an individual
document.
1. words tend to come in bunches that appear together, being highly correlated
with some words, but uncorrelated with most.

Taken together, this suggests that we don't need to keep around all the counts
of all the words from all the documents, but that we can maintain most of the
relevant information by **compressing** the representation, projecting onto
a small number of variables. This is exactly what PCA does.

The `nyt_raw` dataset contains some news stories taken from the New York
Times Annotated Corpus (Sandhaus, 2008), which consists of about 1.8 million
stories from the Times, from 1987 to 2007, which have been hand-annotated by
actual human beings with standardized machine-readable information about their
contents. From this corpus, Prof. Shalizi has randomly selected 57 stories about
art and 45 stories about music, and turned them into a bag-of-words data frame,
one row per story, one column per word; plus an indicator in the first column of
whether the story is about art or music. We'll use this indicator to evaluate
how well our projections are working, but one could of course use it to train
a classifier instead.

There are actually 2 datasets in the course package, the first contains the
counts of each word (as illustrated above), while the second, `nyt_tfidf`
has been
renormalized using Term-Frequency, Inverse Document-Frequency weights (TF-IDF
in this context). This maintains the sparsity, while avoiding the need for
scaling. And just as importantly, it deemphasizes high-frequency words that
appear in most documents, while emphasizing distinctive words that appear
frequently in a few documents.

**Q1** Each of the datasets mentioned above occupy about
`r format(object.size(nyt_raw))` of memory on your machine. While they are both
"sparse", in that many entries are zero, as described above, neither is stored
as a sparse object. True sparse objects store only the nonzero entries. For
example, if stored as a sparse matrix, these would occupy only about 1/8 the
memory. Suppose we actually had access to the entire corpus. This would be
1.8 million articles and about 50,000 words. Now imagine "centering and scaling"
this matrix (as usually recommended before performing PCA). Is this a good
idea?

::: {.solbox data-latex=""}
Your answer
:::


**Q2** Perform PCA on `nyt_raw`. You can go ahead and center this data (as
done by default).
Which 20 words have the largest magnitude
loadings onto the first principal component?  Explain why. What happens if you
scale the data?


::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::



**Q3** We want to perform PCA on `nyt_tfidf`. Be sure not to centre (see `?prcomp`).
If you try to scale inside of `prcomp()` you'll get an error like:
```
Error in prcomp.default(...) :
  cannot rescale a constant/zero column to unit variance
```
Which words are "constant" (have 0 variance)? Why?

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::



**Q4** Remove the offensive columns and perform PCA on `nyt_tfidf` without and
then with scaling.
Which 20 words have the largest magnitude loadings onto the first principal component in each case? Does scaling help

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::


**Q5** If your goal is to separate music articles from art articles, scaling
may make this harder.
Explain why we should scale our data generally.
Should we _always_ scale the data?

::: {.solbox data-latex=""}
Your answer
:::


**Q6** Let's proceed with the PCA on `nyt_tfidf` _without_ scaling. Look
at the 20 most positive and 20 most negative words on the first component.
What do you notice? What about the second component?

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::


**Q7** Plot the projections of the articles onto PC2 against the projections
onto PC1 and indicate the musical / art articles using
shape or colour. What do you notice?

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::

