---
title: "Lab 01 - Linear Models, Cross Validation"
author: "[Your Name Here]"
date: "Due 19 September at 11pm"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
---

# Instructions



# Brief description


In this lab, we will work on the `prostate` cancer data set.
The (modified) data set has `r nrow(data("prostate", package="ElemStatLearn"))` observations and `r ncol(data("prostate", package="ElemStatLearn")) - 1` variables; the
variable `lpsa` in the 9th column is the response while the rest are
potential predictors. The data set is available in the `{ElemStatLearn}` package and a
description can be found at
<http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/prostate.info.txt>


# Questions

Load the data into `R` and remove the `train` column.

```{r, message=FALSE}
data("prostate", package = "ElemStatLearn")
library(tidyverse)
prostate <- prostate |> select(-train)
```

You can view the first few observations by typing `head(prostate)`. In this
lab, we are interested in two linear models:
a. The **[full model]** with all 8 predictors, and
a. The **[reduced model]** with the 3 predictors most correlated with the response `lpsa`.

## Q1

Which 3 predictors are the most correlated with `lpsa`?

::: {.solbox data-latex=""}
```{r}
# your code
```
:::


## Q2

Without fitting the linear models, can you tell which model fits the data better in the sense of having a smaller residual sum of squares? Why?

::: {.solbox data-latex=""}
Your answer
:::


Next, we estimate the prediction error using the test set approach as
well as cross validation. Enter the following command in `R` to generate a random sequence of group indices (group 1, 2 or 3) for the observations.

**Note: Using another random seed, or not using one, or running the
`sample` command more than once will give you a different sequence than
the one we will use for marking this worksheet --- make sure you get
this right**.

```{r}
set.seed(400)
gpind <- c(sample(rep(1:3, each = 32)), 1)
```

The first 6 elements of `gpind` should read `r gpind[1:6]`.

## Q3

Fit the full and reduced models using observations
**[not]** in group 1, and then obtain predictions for those in
group 1. What are the mean squared prediction errors for the two
models?

**Hint**: The reduced model uses the 3 predictors most correlated
with the response in the training set. Not necessarily the same
as those above that used _all_ the data.

::: {.solbox data-latex=""}
```{r}
# your code
```
:::



## Q4

Does your result in **Q3** contradict with that in **Q2**, and why?

::: {.solbox data-latex=""}
Your answer
:::


## Q5

Repeat the procedure in **Q3** for groups 2 and 3, so as to perform a 3-fold cross validation. What are the overall mean squared prediction errors for the two models?

::: {.solbox data-latex=""}
```{r}
# your code
```
:::


## Q6

Based on these results, which model has a better
predictive ability? Suggest one advantage of the $K$-fold cross
validation ($K>1$) over the leave-one-out cross validation.

::: {.solbox data-latex=""}
Your answer
:::

