---
title: "Lab 03 - Lasso"
author: "[Your Name Here]"
date: "Due 03 October at 11pm"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
---

# Brief description

This week, we will continue to work on the prostate cancer data. Recall that the 9th variable, `lpsa`, is the
response while the rest are potential predictors. the
variable `lpsa` in the 9th column is the response while the rest are
potential predictors. The data set is available in the `{ElemStatLearn}` package and a
description can be found at
<http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/prostate.info.txt>


# Questions

Recall that the ridge regression imposes an $L_{2}$ penalty on the
magnitudes of the coefficients (i.e., the
$\lambda\boldsymbol{\beta}^{\intercal}\boldsymbol{\beta}$ term in the
objective function). The LASSO, on the other hand, uses an $L_{1}$
penalty instead. For centred response $y_{1},\ldots,y_{n}$ and vector of
predictors $\boldsymbol{x}_{1},\ldots,\boldsymbol{x}_{n}$, we have

$$
\hat{\boldsymbol{\beta}}=\underset{\boldsymbol{\beta}}{\arg\min}\left[\sum_{i=1}^{n}(y_{i}-\boldsymbol{x}_{i}^{\intercal}\boldsymbol{\beta})^{2}+\lambda\sum_{j=1}^{p}|\beta_{j}|\right],
$$
for $\boldsymbol{\beta}=(\beta_{1},\ldots,\beta_{p})^{\intercal}$ and
some penalty $\lambda$.



**Q1** Give one reason why you might use LASSO instead of ridge regression.

::: {.solbox data-latex=""}
Your answer
:::


Like last time, read the data into `R` using the following command (with
correct file path and name):

```{r, message=FALSE}
data(prostate, package = "ElemStatLearn")
library(tidyverse)
prostate <- prostate |> select(-train)
library(glmnet)
```

Again, we use the function `glmnet()` in the package of the same name.
Read the
documentation of the function by typing `?glmnet` in the console. We are
interested in these three arguments of the function: `x`, `y`, `alpha`.

**Q2** Fit the LASSO using the `glmnet()` function. Write down
the code you used for the fitting (just the line involving the
`glmnet()` function).

**Hint**: Recall the `as.matrix` function you used on the design
matrix last time.

::: {.solbox data-latex=""}
```{r}
# your code
```
:::


**Q3** Inspect how the coefficients evolve as the penalty is
relaxed by plotting the fitted object using the `plot()` function.
Write down the first three variables that enter the model.

**Hint**: `plot()` is a method. When invoked, it calls the
plotting function that corresponds to the class of the object
supplied. The relevant documentation in this case is
`?plot.glmnet`.


::: {.solbox data-latex=""}
```{r}
# your code
```
:::


Now, we choose the optimal value of $\lambda$ via cross validation.
Recall how you used `cv.glmnet()` last time with the `nfolds` argument in
addition to those above (check the documentation of this function).

Carry out a 5-fold cross validation. Run
`set.seed(400)` **immediately before** the `cv.glmnet()`
function to obtain reproducible results.

**Q4** What is the value of $\lambda$ that gives the smallest mean squared error?

::: {.solbox data-latex=""}
```{r}
# your code
```
:::


**Q5** How many predictors are in the model with this value of $\lambda$?

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::


We compare this and the optimal ridge regression model obtained last
time. For design matrix $\boldsymbol{X}$ (without intercept) with
singular value decomposition
$\boldsymbol{X}=\boldsymbol{U}\boldsymbol{\Lambda}\boldsymbol{V}$, where
$\boldsymbol{\Lambda}$ is a diagonal matrix of singular values
$d_{1},\ldots,d_{p}$, the degrees of freedom  of the
ridge regression model is
$$\mbox{df}=\sum_{i=1}^{p}\left(\frac{d_{i}^{2}}{d_{i}^{2}+\lambda}\right),$$
where $\lambda$ is the penalty.

**Q6** The optimal ridge regression model (from Lab 2) has
$\lambda=e^{-2.4}$. What is the $df$ of this model?

**Hint**: Centre the $\boldsymbol{X}$ matrix first. You will find
the `svd` function useful. Make sure there's an intercept!!


::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::


**Q7**
From the results in **Q5** and **Q6**, which model is
more complex based on their $df$'s (note that for lasso, this is only an
unbiased estimate of $df$)?

::: {.solbox data-latex=""}
Your answer
:::


