---
title: "Lab 09 - Boosting"
author: "[Your Name Here]"
date: "Due 28 November at 11pm"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
---


# Questions


In this lab we explore the power of the boosting algorithm and also try
to find out when this method could fail.

## Boosting with a "good" data set

Load `goodboost` in the course package. We will try to predict `class`
using the `x` and `y` coordinates. You will need the
`{adabag}` package for the functions below.

```{r setup, message=FALSE, warning=FALSE}
library(adabag)
library(ggplot2)
library(Stat406)
data("goodboost")
data("badboost")
goodboost <- as.data.frame(goodboost) # boosting() won't work with tibbles...
badboost <- as.data.frame(badboost)
library(ggplot2)
theme_set(theme_bw())
ggplot(goodboost, aes(x, y, colour = class)) +
  geom_point() + scale_colour_viridis_d()
```

We set the classifiers $T_{j}$ to be **stumps** --- a stump
is a tree with only one split. To specify this, run the following line:

```{r}
stump <- rpart.control(cp = -1, maxdepth = 1, minsplit = 0, xval = 0)
```

**Q1** Using the function `boosting()`, perform boosting on this data set with
150 trees (iterations). What is the misclassification error rate? Display the confusion matrix.

**Important notes --- read before you fit!**

1. In this part set `boo = FALSE` so that the original sample is used in all trees.
1. Make use of the `control` parameter to incorporate the `stump` settings you specified above.
1. Run the line `set.seed(123)` right before running the boosting algorithm in R.

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::


**Q2** Identify and plot the misclassified observations for the
boosting algorithm up to the $i$th tree, where $i=1, 10, 25,50,100$ and
$150$. You may use the provided `plot_boost()` function to generate your plots. **In the output, include the plot for $i=150$ only.**

**Hint**: The `predict()` function in this package requires the
vector of responses (as a factor) to be present. Which parameter of `predict`
controls the number of iterations to use?

```{r plot-boost}
#'  Plot the results of a fitted boosting object
#'
#'  This function plots the class and classification error of a fitted boosting
#'  object to the boston housing data
#'
#'  @param boost fitted boosting object
#'  @param dat data used to fit the boosting object
#'  @param ntree the number of trees of the boosting object to be used in the prediction
plot_boost <- function(boost, dat, ntree = length(boost$tree)) {
  pr <- predict(boost, newdata = dat, newmfinal = ntree)
  dat$pred <- pr$class
  dat$error <- factor(ifelse(dat$pred == dat$class, "correct", "wrong"),
                      levels = c("correct", "wrong"))
  ggplot(data = dat) +
    geom_point(aes(x, y, colour = pred, shape = error), size = 2) +
    scale_shape_manual(values = c(16, 4), drop = FALSE) +
    scale_colour_viridis_d() +
    theme_bw()
}
```


::: {.solbox data-latex=""}
```{r}
# your code
```
:::


## Boosting with a "bad" data set


**Q3** Repeat **Q1** and **Q2** using the
`bad_boost` data. What is the misclassification error rate? Does
increasing the number of iterations help? **Be sure to set the seed again.**

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::



**Q4** Now allow the trees to grow a bit deeper by setting
`maxdepth = 2` in the `rpart.control` command while keeping other
parameters fixed at those in **Q1**. Let the algorithm iterate for
100 times. What do you observe?

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::



## What's the difference?

The response variable in the data set `badboost` was
generated through the following relationship:
$$\frac{1}{2}\log\left(\frac{\mathbb{P}\left(Y=a\ |\ \boldsymbol{X}=\boldsymbol{x}\right)}{1-\mathbb{P}\left(Y=a\ |\ \boldsymbol{X}=\boldsymbol{x}\right)}\right)=\left[(x_{2}-2)_{+}-(x_{1}+1)_{+}\right](1-x_{1}+x_{2})$$
where $(z)_{+}=\max(z,0)$. Recall that we attempt to estimate
$$\frac{1}{2}\log\left(\frac{\mathbb{P}\left(Y=a\ |\ \boldsymbol{X}=\boldsymbol{x}\right)}{1-\mathbb{P}\left(Y=a\ |\ \boldsymbol{X}=\boldsymbol{x}\right)}\right)$$
using a **linear combination** of the individual classifiers
$T_{j}(\boldsymbol{x})$ in boosting.

**Q5** We observed that boosting stumps yields good results on
the first data set but not the second. Why is it so?

::: {.solbox data-latex=""}
Your answer
:::


