---
title: "Lab 06 - Generative Models"
author: "[Your Name Here]"
date: "Due 31 October at 11pm"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
---


In this lab we will continue our investigation of the `penguins` data set
from the previous lab.

Here, we'll reduce the problem down to a binary classification problem
by getting rid of the Chinstrap species.

```{r data-processing, message=FALSE, warning=FALSE}
library(MASS)
library(tidyverse)
theme_set(theme_bw())
data("penguins", package = "palmerpenguins")
penguins <- penguins |>
  dplyr::select(species, bill_length_mm, flipper_length_mm) |>
  filter(species != "Chinstrap")
```

- Previously, we explored this dataset using classification trees, which are a **discriminative model**.
- We will now explore this same dataset using a **generative model**:


# A Simple Generative Model: Linear Discriminant Analysis (LDA)

Let $Y \in \{ "Adelie", "Gentoo" \}$ be the species response variable,
and let $X \in \mathbb{R}^2$ be the vector of bill length/flipper length predictors.
The (generative) family of distributions that we will consider are:

$$
P(X=x|Y=y) = \mathcal{N}(x; \mu_y, \Sigma), \qquad \mu_y \in \mathbb{R}^2, \quad \Sigma \in \mathbb{R}^{2 \times 2}.
$$

In other words, we assume that all predictors $X$ for a given species $Y=y$ are normally distributed.
This family of distributions is defined by the following parameters:
- $\mu_\mathrm{Adelie}, \mu_\mathrm{Gentoo} \in \mathbb{R}^2$:
 (the mean of the Normal distribution, which is *specific* to each species), and
- $\Sigma \in \mathbb{R}^{2 \times 2}$: (the covariance of Normal distribution, which is *shared* across all species).


## Q1: Data Simulation

Before we fit this model to the penguins data, let's simulate some data from a distribution in this model.
Assume that the parameters of the distribution are:

- $\mu_\mathrm{Adelie} = (180, 40)$
- $\mu_\mathrm{Gentoo} = (210, 45)$
- $\Sigma = \begin{pmatrix} 100 & 5 \\ 5 & 10 \end{pmatrix}$

Draw 200 representative "penguins" of each species from this distribution,
and plot the resulting data points in a scatterplot, colored by species.

::: {.callout-info}
See `? mvrnorm` for sampling from multivariate normal distributions.
:::

::: {.solbox data-latex=""}
```{r simulation}
# your code
```
:::


**Q2** Now plot the penguins data.
Even though the parameters from the previous question probably don't fit the data,
do you think this family of generative distributions will describe the penguins data well?

::: {.solbox data-latex=""}
```{r penguins-plot}
# your code
```

Your answer
:::



**Q3**
Let's say our goal is still prediction;
i.e. if we observe a new penguin $X = \{ \mathrm{flipper\_len, bill\_len} \}$,
we are interested in predicting its **most probable** species
$Y \in \{ "Adelie", "Gentoo" \}$.

Given a fit LDA model $P(X=x \mid Y=\mathrm{Adelie}), P(X=x \mid Y = \mathrm{Gentoo})$,
how would you use this model to predict the species of a new penguin $X=x$?
What additional information would you need, and how could you estimate it from the training data?
Answer using equations and 3-4 sentences.

::: {.solbox data-latex=""}
Your answer
:::



**Q4**
The following code fits an LDA model with maximum likelihood estimation,
and plots the most probable class predictions
$\mathrm{argmax}_y P_\mathrm{MLE}(Y = y | X = x)$
for each point in space.

```{r q2-solution, message = FALSE, warning=FALSE, fig.width=6, fig.height=3}
ldaobj <- lda(species ~ ., data = penguins)
fine_grid <- map(penguins[, -1], ~ range(.x, na.rm = TRUE)) |>
  map(~ seq(.x[1], .x[2], length.out = 100))

fine_grid <- expand_grid(!!!fine_grid)
fine_grid$pred_class <- predict(ldaobj, newdata = fine_grid)$class

# Plot contours
ggplot(penguins, aes(flipper_length_mm, bill_length_mm)) +
  geom_raster(data = fine_grid, aes(fill = pred_class), alpha = .3) +
  geom_point(aes(colour = species)) +
  scale_fill_viridis_d() +
  scale_colour_viridis_d()
```

Suppose we change the flipper and bill lengths of two
observations from each class (see the figure for the ones that moved.
Running LDA on this modified data set yields the
following predictions at every point in space.

```{r echo = FALSE, warning=FALSE, fig.width=6, fig.height=3}
extras <- tribble(
    ~ flipper_length_mm, ~ bill_length_mm, ~ species,
    172, 59, "Gentoo",
    173, 58, "Gentoo",
    171, 52, "Adelie",
    172, 50, "Adelie",
)
penguins_extra <- bind_rows(penguins, extras)
dump <- c(186, 254, 99, 130)
penguins_replaced <- penguins_extra[-dump, ]
ldaobj2 <- lda(species ~ ., data = penguins_replaced)
fine_grid$pred_class2 <- predict(ldaobj2, newdata = fine_grid)$class


ggplot(penguins_extra, aes(flipper_length_mm, bill_length_mm)) +
  geom_raster(data = fine_grid, aes(fill = pred_class2), alpha = .3) +
  geom_point(aes(colour = species)) +
  scale_colour_viridis_d() +
  scale_fill_viridis_d() +
  annotate(
    geom = "segment",
    arrow = arrow(),
    color = "orange",
    x = penguins_extra$flipper_length_mm[dump],
    y = penguins_extra$bill_length_mm[dump],
    xend = extras$flipper_length_mm,
    yend = extras$bill_length_mm)
```

Note that the **decision boundary** - i.e. the line in $X$-space where the most probable class prediction changes -
is quite different, even though we've only changed 4 points. Why? (Answer in 2-3 sentences.)

::: {.solbox data-latex=""}
Your answer
:::


**Q5**: Now let's imagine we had used a discriminative model, like logistic regression, to classify the penguins.
Do you think this model would have been as sensitive to the change in these few points?
Answer in 1-2 sentences.

::: {.solbox data-latex=""}
Your answer
:::

