---
title: "Lab 04 - Basis Regression"
author: "[Your Name Here]"
date: "Due 10 October at 11pm"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
---

# Brief description


This week, we will perform spline regression using B-splines. The
training and test data sets contain the head acceleration
(`accel`, in *g*) as a function of time (`times`, in ms) after impact in
a simulated motorcycle accident. The original data set is in the `MASS`
package. A description can be found by typing `?mcycle`.

Questions
=========

In spline regression, the number of knots $K$ (i.e. number of basis functions) controls how adaptive the
regression line is. We will compare three fits with $K=5$, 10 and 20.
First, read the training data using the following statement:

```{r, message = FALSE}
library(Stat406)
library(tidyverse)
```

Then, create a vector of 5 knots using the following function:

```{r}
knot_maker <- function(k) quantile(mcycle_train$times, (1:k) / (k + 1))
```
Use similar statements to create the vectors for $K=10$ and 20.

::: {.solbox data-latex=""}
```{r knots}
k <- c(5, 10, 20)
knots <- map(k, knot_maker)
```
:::


**Q1**
Explain how are the locations of the knots determined.
Are they equally spaced along the *x*-coordinate?

::: {.solbox data-latex=""}
Your answer
:::


Fit the cubic splines using the B-spline basis (`bs()` function in package
`{splines}`) defined by the knots you created above. Refer to the lectures
on how you can use the `bs()` function inside `lm()`.

**Note 1**: We use cubic splines this time, and thus you do not need to
specify the `degree` argument.

**Note 2**: You need to fit the three splines separately, each using one
set of knots created above.

::: {.solbox data-latex=""}
```{r}
# your code
```
:::


**Q2** Plot the resulting fits. Which one would you choose for
prediction? Explain your choice. **Do not use the test set at this
point**.

**Hint**: Use the `predict()` function to get fitted values, either on
the observed times or a sequence of user-specified times.

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::


**Q3** Now, use the test set to estimate the mean squared
prediction errors of the three regression models. Which model
appears to be the best?

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::


Recall that a smoother is linear if the fitted values can be written as
$$\hat{\boldsymbol{y}}=\boldsymbol{S}\boldsymbol{y},$$
where $\boldsymbol{y}$ is the vector of response with fitted values
$\hat{\boldsymbol{y}}$, and $\boldsymbol{S}$ is a matrix that does not
depend on $\boldsymbol{y}$. For the linear smoother, the effective degrees of freedom is given by $\mbox{tr}(\boldsymbol{S})$.

**Q4** Are the above smoothing splines linear smoothers?

::: {.solbox data-latex=""}
Your answer
:::



**Q5** Is it possible to obtain the effective degrees of
freedom for the three splines you have just fitted? If yes, write
down the answer below. If not, explain why.

::: {.solbox data-latex=""}
Your answer
:::

