---
title: "Lab 02 - Ridge Regression"
author: "[Your Name Here]"
date: "Due 26 September at 11pm"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
---

# Brief description

In this lab, we will perform ridge regression on the prostate cancer data set studied last time. Recall that the 9th variable, `lpsa`, is the response while the rest are potential predictors. the variable `lpsa` in the 9th column is the response while the rest are potential predictors. The data set is available in the `{ElemStatLearn}` package and a description can be found at <http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/prostate.info.txt>

# Questions

Like last time, read the data into R using the following command (with correct file path and name):

```{r, message=FALSE, warning=FALSE}
data(prostate, package = "ElemStatLearn")
library(tidyverse)
prostate <- prostate |> select(-train)
```

## Q1

What are the coefficients for the predictors `lcavol`, `svi` and `lcp` in the full linear regression model? (If you recall, these are the three predictors most correlated with the response.)

::: {.solbox data-latex=""}
```{r}
# Fit the full model with all predictors
fit_full <- lm(lpsa ~ ., data = prostate)

# Look at the coefficients
coef(fit_full)[c("lcavol", "svi", "lcp")]

summary(fit_full)
```
:::

For responses $y_{1},\ldots,y_{n}$, vector of predictors $\boldsymbol{x}_{1},\ldots,\boldsymbol{x}_{n}$ and fixed penalization parameter $\lambda$, ridge regression solves the optimization problem $$\hat{\boldsymbol{\beta}}=\underset{\boldsymbol{\beta}}{\arg\min}\left[\sum_{i=1}^{n}(y_{i}-\boldsymbol{x}_{i}^{\intercal}\boldsymbol{\beta})^{2}+\lambda\boldsymbol{\beta}^{\intercal}\boldsymbol{\beta}\right],\label{eq:ridgereg}$$ where $\boldsymbol{\beta}$ is the parameter vector (regression coefficients) and $\lambda$ is a tuning parameter that penalizes $\boldsymbol{\beta}$'s that are "too large". It can be shown that the solution to ridge regression is $$\hat{\boldsymbol{\beta}}=(\boldsymbol{X}^{\intercal}\boldsymbol{X}+\lambda\boldsymbol{I})^{-1}\boldsymbol{X}^{\intercal}\boldsymbol{y},\label{eq:estpar}$$ where $\boldsymbol{X}$ is the design matrix, $\boldsymbol{y}$ is the response vector and $\boldsymbol{I}$ is the identity matrix of appropriate size.

## Q2

What is the value of $\lambda$ for the regression considered in **Q1**? As $\lambda\to\infty$, what is the behaviour of the vector $\hat{\boldsymbol{\beta}}$ (the ridge regression estimator)?\

::: {.solbox data-latex=""}
Your answer
:::

Now, we perform ridge regression using the function `glmnet` in the package of the same name.Load the package using `library(glmnet)`. Read the documentation of the function by typing `?glmnet` in the console. We are interested in these four arguments of the function: `x`, `y`, `alpha` and `lambda`.

We consider the sequence of penalties $\boldsymbol{\lambda}=(e^{-3},e^{-2.8},e^{-2.6},\ldots,e^{2.6},e^{2.8},e^{3})$. Create this vector of length 31 and store it as variable `lam`. Fit the series of ridge regressions to the data set. You will need to use the following command:

```{r, eval = FALSE}
fittedobject <- glmnet(x = ___, y = ___, alpha = 0, lambda = ___)
```

where `alpha = 0` specifies the ridge regression (as opposed to other shrinkage methods). Replace the `___` by the appropriate variable names.

**Hint**: The argument `x` accepts a matrix. You can use `as.matrix()` to convert a data frame to matrix. Or, combine a formula with `model.matrix()` like `model.matrix(y ~ var1 + var2, data = df)`. Try `?model.matrix`. Note that `model.matrix()` adds an intercept column, which `glmnet()` does as well, so you'll have to get rid of one if you go this route.

::: {.solbox data-latex=""}
```{r}
library(glmnet)

# 1) Lambda sequence: e^{-3}, e^{-2.8}, ..., e^{2.8}, e^{3}  (length = 31)
lam <- exp(seq(-3, 3, by = 0.2))

# 2) Design matrix (no intercept column because glmnet adds it)
X <- model.matrix(lpsa ~ ., data = prostate)[, -1]

# 3) Response
y <- prostate$lpsa

# 4) Fit ridge regressions across the lambda grid
fittedobject <- glmnet(x = X, y = y, alpha = 0, lambda = lam)

# (Optional) peek
# fittedobject
coef(fittedobject, s = lam[1])     # coefficients at the first lambda
coef(fittedobject, s = lam[length(lam)])  # at the last lambda

```
:::

## Q3

Notice that the fitted `glmnet` object is a list with many components. Inspect it by looking in the Environmet tab, printing it, calling `names(obj)` and / or trying `str(obj)`. Like most modelling routines in `R`, there is a `coef()` method for `glmnet` objects.

Write down the coefficients for the predictors `lcavol`, `svi` and `lcp` when $\lambda=e^{3}$ and $\lambda=e^{0}$, respectively.

**Hint**: `fittedobject` is a list. Variables in a list can be extracted using the `$` operator. For example, `fittedobject$lambda` gives you the $\lambda$ values used in the series of ridge regressions.

::: {.solbox data-latex=""}
```{r}
# coefficients at lambda = exp(3)
coef_exp3 <- coef(fittedobject, s = exp(3))
coef_exp3[c("lcavol", "svi", "lcp"), ]

# coefficients at lambda = exp(0) = 1
coef_exp0 <- coef(fittedobject, s = exp(0))
coef_exp0[c("lcavol", "svi", "lcp"), ]

```
:::

Next, we find the optimal value of $\lambda$ (that yields the best predictive ability) via cross validation. The relevant function is `cv.glmnet()`. We will also need to use the `nfolds` argument in addition to those above (check the documentation of this function).

## Q4

Carry out a 5-fold cross validation, using the same $\lambda$ values as above. Run `set.seed(406)` **immediately before** the `cv.glmnet()` function to obtain reproducible results. Inspect the fitted object; which variable stores the cross validation mean squared errors?

::: {.solbox data-latex=""}
```{r}
library(glmnet)

# same lambda grid as before
lam <- exp(seq(-3, 3, by = 0.2))
X <- model.matrix(lpsa ~ ., data = prostate)[, -1]
y <- prostate$lpsa

set.seed(406)
cvfit <- cv.glmnet(
  x = X, y = y,
  alpha = 0,
  lambda = lam,
  nfolds = 5,
  type.measure = "mse"   # explicit: MSE for gaussian
)

# The cross-validated mean squared errors (one per lambda):
cv_mse <- cvfit$cvm
head(cv_mse)

```

Your answer
:::

## Q5

What is the value of $\lambda$ that minimizes the CV error and what is its associated mean squared error?

**Hint**: You can visualize the result using `plot()` on the fitted object in **Q4**.

::: {.solbox data-latex=""}
```{r}
# lambda that minimizes CV error
best_lambda <- cvfit$lambda.min

# associated CV mean squared error
best_mse <- min(cvfit$cvm)

best_lambda
best_mse

plot(cvfit)


```
:::

## Q6

For this (optimal) $\lambda$, what are the fitted coefficients for the predictors `lcavol`, `svi` and `lcp`?

::: {.solbox data-latex=""}
```{r}
# Coefficients at optimal lambda (lambda.min)
coef_opt <- coef(cvfit, s = cvfit$lambda.min)

# Extract lcavol, svi, and lcp
coef_opt[c("lcavol", "svi", "lcp"), ]


coef(fittedobject, s = cvfit$lambda.min)[c("lcavol", "svi", "lcp"), ]

```
:::
