---
title: "Lab 05 - Classification Trees"
author: "[Your Name Here]"
date: "Due 17 October at 11pm"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
---

This week, we will perform LDA on the well-known `palmerpenguins` data set. This data will
be used to classify observations into one of the three penguin species. We'll
use only two of the predictors: `bill_length_mm` and `flipper_length_mm`.

See the website for descriptions of the data and fun pictures of penguins:
<https://allisonhorst.github.io/palmerpenguins/>.

```{r data-processing, message=FALSE, warning=FALSE}
library(tidyverse)
theme_set(theme_bw())
data("penguins", package = "palmerpenguins")
penguins <- penguins |>
  dplyr::select(species, bill_length_mm, flipper_length_mm)
```


# Questions

A classification tree separates the feature space into
(axis-parallel) regions, where the optimal split at each step is
chosen to minimize a measure of homogeneity (Gini index or deviance).
The majority class in each region is the predicted value.

We use the `rpart` function in the package of the same name to fit a
classification tree. Use the control setting
`myc <- rpart.control(minsplit = 3,cp = 1e-8)` and fit the tree using the
following:

```{r}
library(rpart)
myc <- rpart.control(minsplit = 3, cp = .005)
set.seed(800)
```

::: {.solbox data-latex=""}
```{r}
# your code
```
:::


The `control` parameter tells the function to use the settings given by
`myc`, and `method = 'class'` specifies a classification (not regression)
tree. The "`___`" part (model formula) is for you to fill in.

**Q1**
Use the `plot` and `text` functions to draw the fitted
tree with labels. How many terminal nodes are there?

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::



**Q2**
Which **[terminal]** node(s) with at least 10 are the most
homogeneous, i.e., having the most even split between the two
species? Write down the corresponding region of the feature space
and the number of observations in each class within this region.

::: {.solbox data-latex=""}
Your answer
:::


**Q3**
Suppose we have two new observations: $\boldsymbol{x}_{1}$ with `flipper = 173`,
`bill = 43` and $\boldsymbol{x}_{2}$ with `flipper = 196` and `bill = 51`.
What are the predicted species based on this fitted tree?

::: {.solbox data-latex=""}
```{r}
# your code
```
:::


**Q4**

Is the misclassification rate (i.e., the proportion of
observations wrongly classified) a good measure of the predictive
ability of this tree? Why or why not?

::: {.solbox data-latex=""}
Your answer
:::




**Q5**
Find the observations that are misclassified by this
tree. What is the misclassification rate?

**Hint**: Use `predict` with appropriate arguments to obtain the
fitted classes and compare them with the observed response.


::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::


**Q6**
The `printcp()` function displays the cross-validated
error rates (refer to the `xerror` and `xstd` column of the output). Based on
the result, do you think a simpler tree (with fewer splits) may be
more useful for prediction? Why?

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::

