---
title: "Lab 01 - Linear Models, Cross Validation"
author: "[Your Name Here]"
date: "Due 19 September at 11pm"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
editor_options: 
  markdown: 
    wrap: 72
---

# Instructions

# Brief description

In this lab, we will work on the `prostate` cancer data set. The
(modified) data set has
`r nrow(data("prostate", package="ElemStatLearn"))` observations and
`r ncol(data("prostate", package="ElemStatLearn")) - 1` variables; the
variable `lpsa` in the 9th column is the response while the rest are
potential predictors. The data set is available in the `{ElemStatLearn}`
package and a description can be found at
<http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/prostate.info.txt>

# Questions

Load the data into `R` and remove the `train` column.

```{r, message=FALSE}
data("prostate", package = "ElemStatLearn")
library(tidyverse)
prostate <- prostate |> select(-train)
```

You can view the first few observations by typing `head(prostate)`. In
this lab, we are interested in two linear models: a. The **[full
model]** with all 8 predictors, and a. The **[reduced model]** with the
3 predictors most correlated with the response `lpsa`.

## Q1

Which 3 predictors are the most correlated with `lpsa`?

::: {.solbox data-latex=""}
```{r}
# Compute correlations with lpsa
cors <- cor(prostate)[, "lpsa"]

# Sort by absolute correlation (strongest first)
sort(abs(cors), decreasing = TRUE)



prostate %>%
  select(-lpsa) %>%                        # drop response so only predictors remain
  summarise(across(everything(),
                   ~ cor(.x, prostate$lpsa))) %>%  # compute correlation with lpsa
  pivot_longer(everything(),
               names_to = "predictor",
               values_to = "correlation") %>%
  mutate(abs_corr = abs(correlation)) %>%
  arrange(desc(abs_corr)) %>%
  slice_head(n = 3)
```
:::

## Q2

Without fitting the linear models, can you tell which model fits the
data better in the sense of having a smaller residual sum of squares?
Why?

::: {.solbox data-latex=""}
Your answer
:::

Next, we estimate the prediction error using the test set approach as
well as cross validation. Enter the following command in `R` to generate
a random sequence of group indices (group 1, 2 or 3) for the
observations.

**Note: Using another random seed, or not using one, or running the
`sample` command more than once will give you a different sequence than
the one we will use for marking this worksheet --- make sure you get
this right**.

```{r}
set.seed(400)
gpind <- c(sample(rep(1:3, each = 32)), 1)
```

The first 6 elements of `gpind` should read `r gpind[1:6]`.

## Q3

Fit the full and reduced models using observations **[not]** in group 1,
and then obtain predictions for those in group 1. What are the mean
squared prediction errors for the two models?

**Hint**: The reduced model uses the 3 predictors most correlated with
the response in the training set. Not necessarily the same as those
above that used *all* the data.

::: {.solbox data-latex=""}
```{r}
# # Training set = not in group 1
# train <- prostate[gpind != 1, ]
# 
# # Test set = in group 1
# test  <- prostate[gpind == 1, ]

# Add group index as a column
prostate2 <- prostate %>%
  mutate(group = gpind)

# Training = not in group 1
train <- prostate2 %>% filter(group != 1)

# Test = in group 1
test  <- prostate2 %>% filter(group == 1)


# --- Full model (all 8 predictors, exclude 'group') ---
fit_full <- lm(lpsa ~ . - group, data = train)

# --- Reduced model (lcavol, svi, lcp only) ---
fit_reduced <- lm(lpsa ~ lcavol + svi + lcp, data = train)


# --- Predictions on test set (group 1) ---
pred_full <- predict(fit_full, newdata = test)
pred_reduced <- predict(fit_reduced, newdata = test)

# --- Mean Squared Prediction Error ---
mspe_full <- mean((test$lpsa - pred_full)^2)
mspe_reduced <- mean((test$lpsa - pred_reduced)^2)

mspe_full
mspe_reduced
```
:::

## Q4

Does your result in **Q3** contradict with that in **Q2**, and why?

::: {.solbox data-latex=""}
Your answer
:::

## Q5

Repeat the procedure in **Q3** for groups 2 and 3, so as to perform a
3-fold cross validation. What are the overall mean squared prediction
errors for the two models?

::: {.solbox data-latex=""}
```{r}
# Define reduced model predictors (from training set analysis earlier)
reduced_predictors <- c("lcavol", "svi", "lcp")

# Storage for MSPEs
mspe_full <- c()
mspe_reduced <- c()

# Loop over groups 1, 2, 3
for (g in 1:3) {
  # Split into train/test
  train <- prostate2 %>% filter(group != g)
  test  <- prostate2 %>% filter(group == g)
  
  # Fit models
  fit_full <- lm(lpsa ~ . - group, data = train)
  fit_reduced <- lm(
    as.formula(paste("lpsa ~", paste(reduced_predictors, collapse = " + "))),
    data = train
  )
  
  # Predictions
  pred_full <- predict(fit_full, newdata = test)
  pred_reduced <- predict(fit_reduced, newdata = test)
  
  # Compute MSPE for this fold
  mspe_full[g] <- mean((test$lpsa - pred_full)^2)
  mspe_reduced[g] <- mean((test$lpsa - pred_reduced)^2)
}

# Overall mean MSPEs across 3 folds
mean_mspe_full <- mean(mspe_full)
mean_mspe_reduced <- mean(mspe_reduced)

mean_mspe_full
mean_mspe_reduced
```
:::

## Q6

Based on these results, which model has a better predictive ability?
Suggest one advantage of the $K$-fold cross validation ($K>1$) over the
leave-one-out cross validation.

::: {.solbox data-latex=""}
Your answer
:::
