---
title: "Lab 08 - Random Forests and Bagging"
author: "[Your Name Here]"
date: "Due 21 November at 11pm"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
---

# Brief description

In this lab, we will perform Random Forests and Bagging on the `popmusic`
data. See `?popmusic` for information on the available predictors. The
response is `artist`. We're trying to classify the artist using some information
about the individual songs.


```{r setup, warning=FALSE, message=FALSE}
library(randomForest)
library(tidyverse)
library(Stat406)
theme_set(theme_bw())
```

# Questions

A classification tree can be sensitive to the input data. Bagging or
*bootstrap aggregation* tries to produce more stable predictions by
repeatedly fitting trees to bootstrapped samples (i.e., sampling from
the original data set with replacement). The final prediction is the
majority class based on the $K$ trees, where $K$ is the number of
bootstrap samples trained. By
making a single change to bagging --- that in each split of the tree we
restrict the algorithm to choosing among only $m$ randomly selected
features (smaller than $p$, the total number of features), we obtain the
random forest procedure that reduces the correlation among the trees.

We use the `randomForest()` function in the package of the same name to
fit a classification tree. Use the following lines to fit a random
forest to the training data with 1,501 trees:

**Q1** How many features are randomly selected as possible
candidates in each split (i.e., the value of $m$ in the above
description)? **Hint**: See `?randomForest()`.


::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::



**Q2** Plot the fitted object. Do you think using 1,501 trees
    is sufficient? Why?

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::



**Q3** Using the `varImpPlot()` function, write down the 3 most
important variables in terms of their contribution to the decrease
in homogeneity.

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::


**Q4** What is the out-of-bag error rate for the random forest fit in **Q1**?


::: {.solbox data-latex=""}
Your answer
:::


**Q5** What is the value of $m$ (number of candidate variables
in each split) that corresponds to bagging? Using this value of $m$,
perform bagging by including also the argument `mtry = m` in
`randomForest()`. What is the OOB error using bagging?

**[Again, make sure you run `set.seed(200)` before running `randomForest()`.]**

::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::


**Q6** Obtain the misclassification rate based on the test
set, using the random forest fit in **Q1**. Do you expect this value
to be similar to the OOB error rate? Why?


::: {.solbox data-latex=""}
```{r}
# your code
```

Your answer
:::



**Q7** In the random forest fitted in **Q1**, how many
bootstrapped samples (out of 1,501) do **not** contain the third
observation? How many of these trees predict this observation (which
is OOB to them) as `"Radiohead"`?

**Hint**: You can obtain this information from the fitted `randomForest` object.

::: {.solbox data-latex=""}
Your answer
:::



**Q8** Notice that the random forest OOB error is slightly lower than
the bagging OOB error for this data set. Provide two
advantages of random forests over bagging for decision trees.


::: {.solbox data-latex=""}
Your answer
:::

